---
title: "Tornadoes in Cities"
output: html_document
editor_options: 
  chunk_output_type: console
---

What cities (core-based statistical areas--cbsa) have seen the biggest change in tornado frequency before/after 1985? Here I divide the tornado record into two epochs: 1985-2019 and 1950-1984 and then count the number of tornadoes within each core-based statistical area.

Load packages.
```{r}
library(sf)
library(tidyverse)
library(tidycensus)
```

Get USA boundary map from {tidycensus}.
```{r}
data(state_laea)

state_laea <- state_laea |>
  st_transform(crs = 32615)
plot(state_laea)
```

The Core-Based Statistical Areas (CBSA) are geographic locations neighboring urban areas of 10,000+ population and/or are socioeconomically tied to the urban center by commute proximity. Prior to the year 2000, these were collectively known as MSA- Metropolitan Statistical Areas and Micropolitan Statistical Areas. An exclusive feature of GreatData CBSA Data is that the population statistics are calculated annually. Therefore, the CBSA rank is based on a more recent population than typical census data classifications.

https://catalog.data.gov/dataset/tiger-line-shapefile-2019-nation-u-s-current-metropolitan-statistical-area-micropolitan-statist https://www2.census.gov/geo/tiger/TIGER2019/CBSA/tl_2019_us_cbsa.zip

Get the CBSA polygons.
```{r}
if(!"tl_2019_us_cbsa" %in% list.files()){
unzip("tl_2019_us_cbsa.zip",
      exdir = "tl_2019_us_cbsa")
}

CBSA.sf <- read_sf(dsn = "tl_2019_us_cbsa", 
                   layer = "tl_2019_us_cbsa") |>
  st_transform(crs = st_crs(state_laea))
```

Get the tornado tracks. Tracks that recorded at least EF2 damage somewhere.
```{r}
if(!"1950-2019-torn-aspath" %in% list.files()) {
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2019-torn-aspath.zip",
              destfile = "1950-2019-torn-aspath.zip")
unzip("1950-2019-torn-aspath.zip")
}

Torn.sf <- st_read(dsn = "1950-2019-torn-aspath", 
                   layer = "1950-2019-torn-aspath") |>
  st_transform(crs = st_crs(state_laea)) |>
  mutate(Late = yr >= 1985) |>
  filter(mag >= 2)
```

Create a map showing one CBSA and the intersecting tornado tracks. Colors from https://image-color-picker.com/hex-code-picker
```{r}
NameC <- "Atlanta-Sandy Springs-Alpharetta, GA"
Area.sf <- CBSA.sf |>
  filter(NAME == NameC)

State.sf <- state_laea |>
  filter(GEOID == "13")

Intersects <- Area.sf |>
  st_intersects(Torn.sf) |>
  unlist()

AreaTors.sf <- Torn.sf[Intersects, ]

nT <- AreaTors.sf |>
  group_by(Late) |>
  summarize(nT = n()) |>
  pull(nT)

ggplot() +
  geom_sf(data = State.sf, fill = "white") +
  geom_sf(data = Area.sf, color = "gray70") +
  geom_sf(data = AreaTors.sf, 
          mapping = aes(color = Late)) +
  scale_color_manual(values = c("#9E1FDE", "#5FDE1F"),
                     guide = "none") +
  labs(title = paste0("Approximate tracks of significant tornadoes\n", NameC),
       subtitle = paste0("Total numbers: 1950-1984 (purple): ", nT[1], 
                         paste0(", 1985-2019 (green): ", nT[2])))

```

All metro areas. Also see {stars.Rmd} for creating a space-time object.
```{r}
Intersects <- CBSA.sf |>
  st_intersects(Torn.sf) |>
  unlist()
AreaTors.sf <- Torn.sf[Intersects, ]

Intersects.m <- CBSA.sf |>
  st_intersects(Torn.sf, sparse = FALSE)

TorCounts.sf <- 
  data.frame(GEOID = CBSA.sf$GEOID, 
             NAME = CBSA.sf$NAME,
             nT = rowSums(Intersects.m),
             geometry = CBSA.sf$geometry) |>
  st_as_sf()
```

```{r}
TorCounts.sf |>
  arrange(desc(nT))

TorCounts.sf |>
  slice_max(nT, n = 25) |>
  ggplot(mapping = aes(y = reorder(NAME, nT), x = nT)) +
  geom_point(size = 2, colour = "black") + 
  geom_segment(mapping = aes(yend = NAME, xend = 0), size = 1)+
  labs(y = "", x = "") +
  theme_minimal()
```

```{r}
library(patchwork)

Top25.sf <- TorCounts.sf |>
  slice_max(nT, n = 25)

( US_map <- ggplot(data = state_laea) +
  geom_sf(fill = 'gray90', color = 'white', size = .15) +
  geom_sf(data = Top25.sf, size = 0,
          mapping = aes(fill = nT)) + 
  scale_fill_distiller(palette = "Oranges",
                       direction = 1, 
                       guide = 'none') + 
  theme_void() 
)

( P_plot <- ggplot(data = Top25.sf, 
                   mapping = aes(x = nT, y = reorder(NAME, nT), color = nT)) +
  geom_point(size = 2) +
  scale_color_distiller(palette = "Oranges",
                        direction = 1, 
                        guide = 'none') + 
  scale_x_continuous(limits = c(0, NA)) +
  labs(title = "Top 25 Metropolitan Areas by Number of Significant Tornadoes",
       subtitle = "1950-2019",
       y = "", x = "") + 
  theme_minimal() +
    theme(axis.ticks = element_blank(),
          panel.grid  = element_blank()) )

P_plot / US_map
```

Create a link/hover with map. https://twitter.com/kyle_e_walker/status/1397530708218892293 https://gist.github.com/walkerke/3628171efae66421c299c9f2dbee0f34

```{r}
library(ggiraph)

( US_map <- ggplot(data = state_laea) +
  geom_sf(fill = 'gray30') +
  geom_sf_interactive(data = Top25.sf, size = 0,
                      mapping = aes(fill = nT, data_id = GEOID)) + 
  scale_fill_distiller(palette = "Greens",
                       direction = 1, 
                       guide = 'none') + 
  theme_dark() )

( P_plot <- ggplot(data = Top25.sf, 
                   mapping = aes(x = nT, y = reorder(NAME, nT), color = nT)) +
  geom_point_interactive(size = 2, mapping = aes(data_id = GEOID)) +
  scale_color_distiller(palette = "Greens",
                        direction = 1, 
                        guide = 'none') +
  scale_x_continuous(limits = c(0, NA)) +
  labs(title = "Top 25 Metropolitan Areas by Number of Significant Tornadoes",
       subtitle = "1950-2019",
       y = "", x = "") + 
  theme_dark() )

girafe(ggobj = P_plot / US_map, width_svg = 10, height_svg = 6) |>
  girafe_options(opts_hover(css = "fill:cyan;"))
```

Counts grouped early and late
```{r}
Intersects <- CBSA.sf |>
  st_intersects(Torn.sf) |>
  unlist()
AreaTors.sf <- Torn.sf[Intersects, ]

AreaTors.sf$GEOID <- rep(CBSA.sf$GEOID, 
                         times = rowSums(Intersects.m))

TorCounts.df <- AreaTors.sf |>
  st_drop_geometry() |>
  group_by(GEOID, Late) |>
  summarize(nT = n()) |>
  ungroup()

( Diff.df <- TorCounts.df |>
  pivot_wider(names_from = Late, 
              values_from = nT) |>
  rename(After1985 = `TRUE`,
         Before1985 = `FALSE`) |>
  mutate(Difference = After1985 - Before1985,
         DiffPerc = Difference/Before1985 * 100) |>
  drop_na() )

sum(Diff.df$Difference > 0)
sum(Diff.df$Difference < 0)
```

Top 10 and bottom 10 combined.
```{r}
Upward <- Diff.df |>
  slice_max(Difference, n = 10) |>
  mutate(Tendency = "Upward")

Downward <- Diff.df |>
  slice_min(Difference, n = 10) |>
  mutate(Tendency = "Downward")

Both <- rbind(Upward, Downward)

Both <- Both |>
  left_join(CBSA.sf,
            by = "GEOID") 
```

Plot a slope graph.
```{r}
library(ggrepel)

ggplot(Both) +
  geom_segment(mapping = aes(x = .4, xend = .6, 
                             y = Before1985, yend = After1985, 
                             color = Tendency)) +
  scale_color_manual(values = c("#E69F00", "#CC79A7"), guide = 'none') +
  scale_x_continuous(limits = c(0, 1)) +
  theme(panel.background = element_blank(),
        panel.grid=element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank(),
        panel.border=element_blank()) +
  geom_text(data = Both[Both$Tendency == "Upward", ],
            mapping = aes(x = .62, y = After1985, 
                          label = paste(After1985, NAME, sep = "   ")), hjust = 0, size = 3.5, color = "#CC79A7") +
  geom_text(data = Both[Both$Tendency == "Downward", ], 
            mapping = aes(x = .38, y = Before1985, 
                          label = paste(NAME, Before1985, sep = "   ")), hjust = 1, size = 3.5, color = "#E69F00") +
  geom_text(data = Both[1, ], 
            mapping = aes(label = "1950-1984", x = .38, y = 1), hjust = 1, size = 4) +
  geom_text(data = Both[1, ], 
            mapping = aes(label = "1985-2018", x = .62, y = 1), hjust = 0, size = 4) +
  geom_segment(data = Both[1,],
               mapping = aes(x = .45, y = 1, xend = .55, yend = 1), color = "gray70",
               arrow = arrow(angle = 20, length = unit(2, "mm"), type = "closed")) +
  labs(title = "The shifting tornado threat",
       subtitle = "Cities with the largest change in number of significant tornadoes (EF2 or worse damage)",
       x = "", y = "")
```

Create a link/hover with map. https://twitter.com/kyle_e_walker/status/1397530708218892293 https://gist.github.com/walkerke/3628171efae66421c299c9f2dbee0f34

```{r}
Both <- Both |>
  select(GEOID, After1985, Before1985, Difference)

Both.sf <- CBSA.sf |>
  right_join(Both,
            by = "GEOID") 

( US_map <- ggplot(data = state_laea) +
  geom_sf(fill = 'gray30') +
  geom_sf(data = Both.sf, size = 0,
          mapping = aes(fill = Difference > 0)) + 
  scale_fill_manual(values = c("#E69F00", "#CC79A7"),
                    guide = 'none') +
  theme_dark() )

( P_plot <- ggplot(data = Both.sf, 
                   mapping = aes(x = Difference, y = reorder(NAME, Difference), 
                                 color = Difference > 0)) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 0, color = "white") +
  scale_color_manual(values = c("#E69F00", "#CC79A7"),
                      guide = 'none') +
  scale_x_continuous(limits = c(-75, 25)) +
  labs(title = "Top differences in significant tornado counts",
       subtitle = "1985-2019 minus 1950-1984",
       y = "", x = "") + 
  theme_dark() )

P_plot / US_map
```

Get census data from these CBSAs. https://walker-data.com/tidycensus/articles/spatial-data.html
```{r}
options(tigris_use_cache = TRUE)

( Augusta <- get_decennial(geography = "tract",
                           variables = "P001001",
                           year = 2000,
                           state = "GA", 
                           county = "Richmond", 
                           geometry = TRUE,
                           summary_var = "P001001") )
```

